---
title: "Fieldwork App Core Principles"
author: "Nicholas Manna"
date: "7/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction and Resources

This document will cover recurring principles and themes in the Fieldwork App and related apps, such as Porous Pavement, Capture Efficiency, and Inlet Conveyance.
It gives an overview of the back-end and the reactive interactions between user-interface (UI), server, and different modules and tabs. I recommend perusing the following resources before digging into this document: 

[Introduction tutorial](https://shiny.rstudio.com/tutorial/)

[Fieldwork App GitHub Repo](https://github.com/taywater/shiny-fieldwork)

[Modularization](https://shiny.rstudio.com/articles/modules.html)

[PWD RStudio R blog – for RSConnect publishing](http://pwdrstudio/pwd-R-blog/public/)

[Shinyjs – this one’s easy](https://deanattali.com/shinyjs/)

[db Pool (easier than it seems)](https://shiny.rstudio.com/articles/pool-basics.html)

[datatable (DT)](https://rstudio.github.io/DT/)

[reactable](https://glin.github.io/reactable/)

### Core Principle #1: Modules 

The Fieldwork App is composed of modules. The modules represent individual pages or components of the overarching app. App.R brings these components of the app together; calls to `source()` bring the UI and server functions of each constituent module into app.R to be executed. Modules provide a host of benefits for the app's development process. Since modules are self-contained environments, variable names can be re-used between them. Breaking the app into modules also makes expansion, or separation, of the apps much easier since files can be added or removed with minimal adjustments in app.R. 

When modules' source files are brought into app.R, the module functions get called. The modules' UI functions are called within app.R's UI function; likewise their server functions are called within app.R's server function. Global variables can be queried and set in the app.R UI and server functions and passed to modules called there. 

%%%Taylor: These two sentences are unclear - it sounds like there is a difference between the way the app treats globals that are instantiated in the global UI/Server and those instantiated in App.R but not inside those functions. Can you elaborate more on how it treats them differently, perhaps with examples?%%%

These variables are initiated in the functions so they are checked every time the app is loaded. However, libraries, global options, database connections, and file sourcing can be set in the main body of app.R. 

Modules are given distinct namespaces to facilitate the reusability of variable names. These namespaces are created with the function `NS()`, which returns a function that silently modifies variable names within the Shiny environment. The function is instantiated like `ns <- NS(id)`, where `id` is the name of your module. Each input or output for that module must be wrapped in `ns()` to appropriately assign it to the namespace. Creating a module's server function involves calling `moduleServer()`, and `ns` must be called within `moduleServer()`. `moduleServer()` and `ns` work in concert to define and name the environment for the module's code execution.

```{r eval = FALSE}

collection_calendarServer <- function(id, parent_session, ow, deploy, poolConn) {
  
  moduleServer(
    id, 
    function(input, output, session){
      
      #define ns to use in modals
      ns <- session$ns
      
      rv <- reactiveValues()
      
      ...
      #the rest of this server function has been elided for brevity
    }
  )
}

```

Module server functions return outputs that can be passed to other modules as input arguments. This allows modules to communicate with one another, and react to events in modules to which they're connected. This interconnectivity makes it possible for modules to pass control of the application to one another. Fieldwork app modules extensively link to one another in this way. In the example below, module server functions define their outputs and inputs in App.R's server function, allowing for inter-module communication.

``` {r eval = FALSE}

# 2.2: Server Module functions ---------------------------
    # Collection Calendar
    collection_cal <- collection_calendarServer("collection_calendar", parent_session = session,
                                 ow = ow, deploy = deploy, poolConn = poolConn)
    #Add Edit/Location
    ow <- add_owServer("add_ow", parent_session = session, smp_id = smp_id, poolConn = poolConn, deploy = deploy)

```

### Core Principle #2: Table Versions

Tables are queried via a database connection instantiated in App.R. Within the database "mars_testing" there are three schemas: public, fieldwork, and performance. The fieldwork app interacts with data hosted in the fieldwork schema, so `fieldwork.` is often prepended to table names in database queries (`SELECT * FROM fieldwork.active_deployments`). Most tables that are displayed in the app come from views. These views include the base table (ie, deployment), joined with other tables, including lookup tables, so that all the values the user would want to see are already there. 

Tables in the fieldwork app follow a naming convention that describes where the data has come from. Tables directly queried from the databse will have the suffix `_db` appended. When that table is passed through some `dplyr` processing, which may include formatting the date, and reducing and renaming columns, the `_db` will be removed. This new table will be passed to DTOutputs for display. These steps are all reactive, so when an SMP ID changes, the query will update, then the `_db` table will update, the `dplyr` processing will re-run, and the displayed table will update. See the code chunk below.

It's important to have two versions of the table. When a row is clicked, fields are pre-filled based on values from the `_db` table, because not all fields are shown in all tables, and certain values in the displayed table are the lookup description, when we want to enter the lookup UIDs. 

When writing to the database, we write to the original base table, which updates the view on the database side, which the app queries. The app does not react to that, so we add a line to run the query again, which in turn refreshes the displayed table. 

``` {r eval = FALSE}

srt_table_query <- reactive(paste0("SELECT * FROM fieldwork.srt_full WHERE system_id = '", input$system_id, "'"))
rv$srt_table_db <- reactive(odbc::dbGetQuery(poolConn, srt_table_query()))

rv$srt_table <- reactive(rv$srt_table_db() %>% 
                           mutate("test_date" = as.character(test_date), 
                                  "srt_stormsize_in" = round(srt_stormsize_in, 2)) %>% 
                           dplyr::select("system_id", "test_date", "phase", "type", "srt_volume_ft3", "dcia_ft2", "srt_stormsize_in", "srt_summary"))

output$srt_table <- renderDT(
        rv$srt_table(), 
        selection = 'single',
        style = 'bootstrap', 
        class = 'table-responsive, table-hover', 
        colnames = c('System ID', 'Test Date', 'Phase', 'Type', 'Volume (cf)', 'DCIA (sf)', 'Simulated Depth (in)', 'Results Summary') 
      )

```
                                 
 

### Core Principle #4: Editing and Pre-populating Inputs
 
Selecting a row in a site table indicates the need to access more information or edit that row. The Fieldwork App will respond to this indication by modifying the app UI in response to that user's action. When a row is selected, that row's position within the datatable is identified, which is used to reference the row's position and related attributes in the "_db" table. Additionally, rows in other tables are deselected.

Attributes from the selected row (here, `srt_table()$phase`) are then sent into a corresponding UI object (here, the `con_phase` selectInput UI object). Shiny refreshes the UI for the user, and the new value is in place.

``` {r eval = FALSE}

observeEvent(input$srt_table_rows_selected,{ 
  dataTableProxy('future_srt_table') %>% selectRows(NULL)
  updateDateInput(session, "srt_date", value = rv$srt_table()$test_date[input$srt_table_rows_selected])
  
  #update to values from selected row
  updateSelectInput(session, "con_phase", selected = rv$srt_table()$phase[input$srt_table_rows_selected])
  ...
}


```
 
### Core Principle #5: Writing to and Editing the Database

Adding rows to the database, and editing those rows, is triggered by a button click. Different buttons are available for future or completed tests (see CP #3), and different labels are applied based on whether a row is selected. A reactive value for the label is then applied within an `observe`. This conditional is also included within the `observeEvent` for the button click. 

``` {r eval = FALSE}

rv$label <- reactive(if(length(input$srt_table_rows_selected) == 0) "Add New" else "Edit Selected")
observe(updateActionButton(session, "add_srt", label = rv$label()))

rv$future_label <- reactive(if(length(input$future_srt_table_rows_selected) == 0) "Add Future SRT" else "Edit Selected Future SRT")
observe(updateActionButton(session, "future_srt", label = rv$future_label()))

```
 
The Fieldwork App often uses reactive values to manage quotation marks for input strings that will be sent to the database. For instance, when writing to a database, a numeric or text value should be in single quotes, but when writing NULL, there should not be any quotes. Quotes can be set in the reactive value, rather than living inside the `paste()` in the query, where they are permanent and wrap the NA or NULL incorrectly. This also ensures that the blank value for these inputs is always NULL. 

``` {r eval = FALSE} 

rv$test_volume <- reactive(if(is.na(input$test_volume)) "NULL" else paste0("'", input$test_volume, "'"))

```

The following is an example of modal inputs influencing the Fieldwork App's behavior when writing to the database. An SRT is added, or edited, and various choices are made to prepare the inputs to be written depending on the scenario.

Once the "Add/Edit" button is clicked, the conditional is considered to see whether the row is being added or edited. If adding, the app uses an `INSERT INTO` query, which includes the aforementioned reactive values for any input that might be NULL. If editing, the app uses an `UPDATE` query, where srt_uid equals that of the row selected. Since future tests can be selected and edited to be completed tests, check for that and delete the future test. Then, update all affected tables, and clear contents, aside from the root (SMP ID, System ID, Site Name, etc.)

``` {r eval = FALSE}

observeEvent(input$add_srt, {
        if(length(input$srt_table_rows_selected) == 0){
          add_srt_query <- paste0("INSERT INTO fieldwork.srt (system_id, test_date, 
                          con_phase_lookup_uid, srt_type_lookup_uid, 
                          srt_volume_ft3, dcia_ft2, srt_stormsize_in, srt_summary) 
      	                  VALUES ('", input$system_id, "','", input$srt_date, "','", rv$phase(), "', ", 
                                  rv$type(), ",", rv$test_volume(), ",", 
                                  rv$dcia_write(), ", ", rv$storm_size(), ",", rv$srt_summary(), ")")
          ...#other queries and content elided for brevity
          odbc::dbGetQuery(poolConn, add_srt_query)
          
        }else{
          edit_srt_query <- paste0(
            "UPDATE fieldwork.srt SET system_id = '", input$system_id, "', test_date = '", input$srt_date, 
            "', con_phase_lookup_uid = '", rv$phase(),
            "', srt_type_lookup_uid = '",  rv$type(),
            "', srt_volume_ft3 = ", rv$test_volume(),
            ", dcia_ft2 = " , rv$dcia_write(),
            ", srt_stormsize_in = ", rv$storm_size(), 
            ", srt_summary = ", rv$srt_summary(), "
            WHERE srt_uid = '", rv$srt_table_db()[input$srt_table_rows_selected, 1], "'")
          ...
          dbGetQuery(poolConn, edit_srt_query)
        }
          
          #if editing a future test to become a completed test, delete the future test
         if(length(input$future_srt_table_rows_selected) > 0){
          odbc::dbGetQuery(poolConn, paste0("DELETE FROM fieldwork.future_srt 
                                            WHERE future_srt_uid = '", rv$future_srt_table_db()[input$future_srt_table_rows_selected, 1], "'"))
         }
          
          #update srt_table with new srt
        rv$srt_table_db <- reactive(odbc::dbGetQuery(poolConn, srt_table_query()))
        
        #update srt view with new/edited srt
        rv$all_srt_table_db <- reactive(dbGetQuery(poolConn, all_srt_table_query))
        
        #update future srt table in case a future srt was delisted 
        rv$future_srt_table_db <- reactive(odbc::dbGetQuery(poolConn, future_srt_table_query()))
        rv$all_future_srt_table_db <- odbc::dbGetQuery(poolConn, all_future_srt_table_query)
          
        ...
        
        #clear contents aside from System ID
        reset("srt_date")
        reset("con_phase")
        reset("srt_type")
        ...
}
          

```

### Core Principle #N: Root Values and UI Malleability

%%%Taylor: This section is extremely dense. You're describing how modules work from a root value that controls the capabilities/actions of the app depending on the context, but you're not leading with that. I think this would work better as the final core principle, as it draws on a number of others. I will take a stab at writing a stripped-down version of this as a new, final principle.%%%

Each tab has a most-important data variable. These variables, called "root values", have substantial effects on the behavior of the tab. For example, Deploy Sensor is controlled at the root by either the SMP ID field or the Site Name field. When a root value is selected (for example, an SMP ID), the app can query the database and enable/disable UI controls based on the chosen value. It could, for example, show all current/previous deployments at the chosen SMP. It could also enable, disable, or change certain UI elements based on the user's selection. In accordance with CP #4, it could also prepopulate certain UI elements.

Other variables can also control app behavior in this way, but each app's root value controls it most of all. Date fields can tell the app whether a record is a past test or future test, and change app behavior accordingly.


``` {r eval = FALSE}
#toggle state for metadata depending on whether a test date is included
observe(toggleState(id = "flow_data_rec", condition = length(input$srt_date) > 0))
observe(toggleState(id = "water_level_rec", condition = length(input$srt_date) > 0))
observe(toggleState(id = "photos_uploaded", condition = length(input$srt_date) > 0))
observe(toggleState(id = "test_volume", condition = length(input$srt_date) > 0))
...

#toggle state (enable/disable) buttons based on whether system id, test date, and srt type are selected (this is shinyjs)
observe(toggleState(id = "add_srt", condition = nchar(input$system_id) > 0 & length(input$srt_date) > 0 &
                      nchar(input$srt_type) >0 & nchar(input$con_phase) > 0))

#toggle state for future srt
observe(toggleState(id = "future_srt", condition = nchar(input$system_id) > 0))

```

###Core Principle N+1: Main Tables and Site Tables

Modules that are composed of several tabs often have tables that show broad expanses of data, and narrow subsets of data (filtered by a root value, in accordance with CP #N). These tables can be considered "main" tables and, because root values often target single sites, "site" tables.

Main tables give an overview of an entire dataset. For example, all completed or future SRTs, regardless of which sites they are associated with, would be considered a main table. In accordance with CP #2, they are queried from database views, minimally modified, and are filtered to produce site tables when certain conditions are met. Values from these site tables (for example, a single selected test record), can be prepopulated into the UI in accordance with CP#4

Site tables display a subset of information from these main tables, filtered in accordance with CP #2. In the example below, an SRT main table is filtered into a site table based on a selection of an system ID, the root value for SRT tests.

In the CP #2 code chunk, you can see that `selection = "single"`. This enables reactions based on a row selection, and only allows one row to be selected at a time; the main table for SRT is a reactable and not a datatable, but has the same argument. First, system ID is selected for the "Add/Edit" tab based on row selection in the main table. Then, the tab switches, and future table is deselected. This is also a recurring trend; to make sure that there is no mix-up, only table of each field test can be selected at a time. Since these queries take time, a delay of 300ms is initiated. The row in the "Add/Edit" system table is selected where the srt_uid equals the srt_uid of the row selected in the main table.

``` {r eval = FALSE}

      #click a row in the all srt table, switch tabs, and select the correct SMP ID, then select the correct test
      observeEvent(input$srt_selected, {
        updateSelectizeInput(session, "system_id", choices = sys_id, 
                             selected = rv$all_srt_table()$system_id[input$srt_selected], 
                             server = TRUE)
        updateTabsetPanel(session = parent_session, "inTabset", selected = "srt_tab")
        #deselect future rows
        updateReactable("all_future_srt_table", selected = NA)
        #delay so that the selectizeInput is updated and table is queried before it is searched by R
        #basically make sure things happen in the right order
        #delay time based on trial and error
        delay(300,{
                srt_row <- which(rv$srt_table_db()$srt_uid == rv$all_srt_table_db()$srt_uid[input$srt_selected], arr.ind = TRUE)
                dataTableProxy('srt_table') %>% selectRows(srt_row)
              }
        )
      })

```

When the database is updated, queries are re-run so they displayed tables also update.
